---
title: "Medic Masters - Intro R"
author: "Katherine Tansey"
date: "7/3/2017"
output:
  html_document:
    depth: 4
    keep_md: yes
    theme: united
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
---


# Data Types
Not all data are the same. R treats data differently depending on what type of data it is. To determine what type of data something is you can use the **class()** function. Variable inherit their class from the value or object assigned to that variable. 

```{r}
class(x)
class(y)
```

# Coercion
Sometime R gets it wrong. Each column of a data frame must be a single data type and this is because each column of a data frame is stored as a vector. When joining columns together with two different data types, R will make a choice:     
    
    Character + Anything_Else  ->  Character    
    Number + Integer or Logical  ->  Number    
    Integer + Logical  ->  Integer     

There are commands to force data into a particular type:    

* force to be a number     
    + as.numeric(data$column)    
* force to be a character   
    + as.character(data$column)   
* force to be a vector   
    + as.vector(data$column)   
* force to be a data.frame    
    + as.data.frame(data)   

> Examine coercion

Load babynames from USA dataset
```{r}
#install.packages("babynames")
library(babynames)
```
Make sex a factor
```{r, eval = FALSE}
babynames$sex <- as.factor(babynames$sex)
```
Change sex into a character and a numeric -- Examine the differences    
```{r, eval = FALSE}
babynames$sex1 <- as.character(babynames$sex)
babynames$sex2 <- as.numeric(babynames$sex)
summary(babynames)
```


# Data Structures
There are a variety of different data structures in R. We are only going to cover three of them:
1. Vector
2. Data Frame
3. Factor

For more information on the rest of them, see [here](http://adv-r.had.co.nz/Data-structures.html), [here](http://www.statmethods.net/input/datatypes.html) and [here](https://rstudio-pubs-static.s3.amazonaws.com/90282_db0ce980e867436fa9f4c4f7695edf0f.html#chapter-5---data-structures)

## Vector
Vectors are collections of values all of the same type (meaning characters, numeric, logicals). 

    c() used to combine values into a vector

```{r}
character_vector <- c("Harry Potter", "Ron Weasley", "Hermione Granger", "Neville Longbottom")
numeric_vector <- c(1,2,3,4,5,6)
logical_vector <- c(TRUE, FALSE, FALSE, TRUE)
```

Subsetting, or extracting, individual elements from a vector occurs by using square bracket notation **[ ]** and the number of the elements in the vector that you want. 

*REMEMBER: R counts from 1*

> Can you guess what these will return?  

```{r, results = "hide"}
character_vector[1]
```

```{r results = "hide"}
character_vector[3]
```

```{r results = "hide"}
character_vector[1:3]
```

```{r results = "hide"}
character_vector[c(1,3)]
```

Mathmatical operations can also occur on vectors like they do on numbers and variables. Let's walk through the following code:

Create two vectors. One for money earned in one week on each weekday. The second is money spent on each weekday in that week. 
```{r}
money_earned <- c(85, 14, 67, 33, 54)
money_spent <- c(62, 42, 36, 75, 94)
```

Caluclate the total (net amount) of money for each weekday
```{r}
money_total <- money_earned - money_spent
money_total
```

Find which days in which we earned money (made more money than we spent). Returns a logical, where TRUE is when money_total is greater than 0 and FALSE when it is not.  
```{r}
made_money <- money_total > 0
made_money
```

Extract out just the total money for the days that we earned money.
```{r}
good_days <- money_total[made_money]
good_days
```

Another possible way to do this, can you see what is occuring here?
```{r}
good_days2 <- money_total[made_money > 0]
good_days2
```

> Play with vectors

Merge two vectors together.
```{r}
HP <- c("Harry Potter", "Ron Weasley", "Hermione Granger")
HG <- c("Katniss Everdeen", "Peeta Mellark", "Gale Hawthorne")

Book_PPL <- c(HP, HG)
```


## Factor
Factors store categorical data. Use the fuction **factor()** to convert a vector in a factor. 

```{r}
sex_vector <- c("Male", "Female", "Female", "Male", "Male")
factor_sex_vector <- factor(sex_vector)
factor_sex_vector
summary(factor_sex_vector)
```
**What is summary function doing?**      
    
Factors can be ordered. This aligns with ordinal data, which is categorical but has a set order to it.    

```{r}
age_vector <- c("35-50", "51-69", "35-50", "18-34", "51-69", "18-34", "35-50", "18-34", "51-69", "18-34","35-50", "35-50")
factor_age_vector <- factor(age_vector, ordered=T, levels = c("Under 18", "18-34", "35-50", "51-69", "Over 70"))
factor_age_vector
summary(factor_age_vector)
```

> Create Factors

```{r}
movie_ratings <- c("bad", "okay", "okay", "good", "amazing", "horrible", "bad", "amazing")
```
1. Convert movie_ratings to non-ordered factor vector and print to screen factor_movie_ratings
```

```
2. Convert movie_ratings to ordered factor vector and print to screen factor_movie_ratings
```

```
Which one do you think is more informative?

## Data Frame
Two dimensional object that include both rows and columns. Columns are variables and rows are the observations. Each column can have a different data type. 

    data.frame() used to create a data frame

Use a data set that is in data.frame format which is preloaded into R called "iris". For more information about this dataset, see [here](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/iris.html).

Print the entire dataset to screen
```{r}
iris
```
Print the first 6 rows. Command the same as in UNIX, just syntax slightly different (meaing UNIX does not use () around commands).
```{r}
head(iris)
```
Print the last 6 rows. Command the same as in UNIX, just syntax slightly different.
```{r}
tail(iris)
```
**How do you alter how many lines get printed with head() or tail()?**

    str() prints nice summary of the data.frame including data type of each column   
    
```{r}
str(iris)
```

*REMEMBER: R counts from 1*

### Retrive elements from a data.frame
> Can you guess what these will return?   

```{r, results = "hide"}
iris[,2]
```

```{r, results = "hide"}
iris[4,]
```

```{r, results = "hide"}
iris[,"Species"]
```

```{r, results = "hide"}
iris[1:2,2]
```

```{r, results = "hide"}
iris[2:5,1:2]
```

```{r, results = "hide"}
iris[c(1,4,7), c(1,3)]
```

```{r, results = "hide"}
iris[which(iris$Sepal.Length > 7.5), ]
```
**What do you think the which() command is doing?**    

### Datafame size

    nrow() to find out the number of rows in the dataframe

```{r}
nrow(iris)
```

    dim() returns the number of rows and columns 

```{r}
dim(iris)
```

    colnames() returns the column names of the dataframe 

```{r}
colnames(iris)
```
